---
title: "Template DAV: Lab 1A"
subtitle: "Data Visualization"
author: "your name"
output: html_document
---


---

# Instructions

1. Save the file in one of your folders

2. Insert R code in R chunks

3. Run the code by clicking the green right arrow at the topright of the chunk

- if an assigment was made, the object appears in the Global Environment 
    
- if output was requested, it should appear in the Rmd file  below the R chunk
    
- if an error message appears, adapt the code until it the error meassage dissappears
    
    - if the code is error free, click on the `Knit` button to render the HTML document

4. To run all previous R chunks, click the downwards arrow at the top right of the R chunk

---

The code for the first two R chunks is given.

Install and load the package `tidyverse` (click the green right arrow to check if the code works properly)

```{r}
if('tidyverse' %in% installed.packages() == FALSE) install.packages('tidyverse')
library(tidyverse)
```


---

The data set for the exercises is `txhousing` of the package `tidyverse`


# `dplyr` functions


a. __Display the variables `volume` and `inventory` using the function `select()`. Next, display all other variables by deselecting `volume` and `inventory` in the function `select()`.__

(First run the code with the green right arrow. The output should appear in your Rmd document below the chunk. If there are no error mesaages, click the `Knit` button to render the HTML file)

```{r}
select(txhousing, volume, inventory)
select(txhousing, -volume, -inventory)
```


If rendering the HTML file work properly, then continue with the exercises. Remember to test each chuck for errors before knitting the HTML file!

---


## `filter()`

The function `filter(data=<DATA>,  <LOGICAL EXPRESSION>)` selects rows of data based on a logical expression. 

b. __Display cases that had less than 10 sales in the year 2011__


```{r}

```

---


## `mutate()`

The function `mutate(data=<DATA>, <NEW_VAR = EXPRESSION>)` computes a new variable and adds it to the data.

c. __Compute the logarithm of the variable `volume` unde  the name `log_volume`. Display the new data set without saving it.__


```{r}

```

---

## `summarise()`

The function `summarise(data=<DATA>, <STAT = function(<VARIABLE>))` computes summary statistics with functions like `mean()` and `sd()`, or  `n()` (see the section Summary Functions on the `dplyr` cheat sheet). `STAT` is a user-supplied name for the summary statistic.

c. __Display the means and standard deviations of the variables `volume` and `sales`. Supply appropriate names.__



```{r}

```

---

## The pipe operator 



h. __Select cases with `sales > 450`, compute the variable `log_sales` as logarithm of `sales`, and display the means (M), standard deviations (SD) and number of observations (Freq) of `log_sales`.__


```{r}

```

---

## `group_by ()`


The function `group_by(data=<DATA>, <VARIABLE(S)>)` groups the data by the value(s) of one or more variables [r4ds: Grouped summaries with `summarise()`](https://r4ds.had.co.nz/transform.html#grouped-summaries-with-summarise). It is especially useful in combination  with `summarise()`

i. __Repeat the previous exercise, but now display the means, standard deviations and number of observations per city.__


```{r}

```

---

## `arrange()`

The function `arrange(data=<DATA>, <VARIABLE(S)>)` sorts the rows of the data by variable(s) in `<VARIABLE>` [r4ds: Arrange rows with `arrange()`](https://r4ds.had.co.nz/transform.html#arrange-rows-with-arrange). It is also useful in combination with `summarise()`.

j. __Repeat the previous exercise, but now sort the output on the means of the cities.__


```{r}

```

---




# `ggplot2` function

Use the [Data Visualization Cheat Sheet](https://github.com/rstudio/cheatsheets/raw/master/data-visualization-2.1.pdf) and the [Tidyverse ggplot2 Reference](https://ggplot2.tidyverse.org/reference/) as a reference for the exercises, and check [r4ds: Data visualisation](https://r4ds.had.co.nz/data-visualisation.html) for examples of `ggplot2` code. 


In this lab we use the dataset `txhousing`. Check the help page for the content of this data set.


## Data

The basic function for making plots with `ggplot2` is  

```
ggplot(data = <DATA>) 
```

where `<DATA>` is the name of the data set. 


a. __Display the plot that is produced by the function `ggplot()` with the `data = txhousing` argument. __


```{r}

```

---


## Aesthetics


Without an aesthetics specification the plot is empty. To assign values to the axes a `mapping` argument is required in which the aesthetics are specified. The generic code for that   

```
ggplot(data = <DATA>, mapping = aes(x = <VARIABLE>, y = <VARIABLE>, . . .)) 
```

where the aesthetic function `aes()` tells `ggplot()` which variables to map to the x- and y-axis. The dots mean that there are other aesthetics, like `alpha` , `color` and `fill`. We will encounter examples of these aesthetics later in the lab. 

b. __Run the `ggplot()` function again, but now with `aes()` specified such that `volume` is mapped to the x-axis and `sales` to the y-axis. __


```{r}

```

---
 



## Geoms

The axes now have values, but nothing is plotted. The `<GEOM_FUNCTION>` determines what is to be plotted, e.g. `geom_point()` for points, `geom_smooth()` for smooth regression lines, `geom_histogram()` for a histogram, etc. It is possible to add more than one geom to a plot.

There are two ways to specify the aesthetics in combination with a geom. When specified in the `ggplot()` function like

```
ggplot(data = <DATA>, mapping = aes(x = <VARIABLE>, y = <VARIABLE>)) + 
   <GEOM_FUNCTION>()
```

the mappings are the default for all following layers. However, if specified as  

```
ggplot(data = <DATA>) + 
   <GEOM_FUNCTION>(mapping = aes(x = <VARIABLE>, y = <VARIABLE>))
```

the mappings only refer to that specific geom.

---

We will now make some plots of the variable `sales`. With a single variable the `aes()` function only requires one argument, either `x` or `y`. 

a. __Display a histogram of `sales` with `geom_histogram()`. Do you map `sales` to the x- or y-axis?.__


```{r}

```

The plot is accompanied by a suggestion and a warning. The default number of bins is 30, but a smaller number woud be  more appropriate. The variable `sales` is also highly skewed, so for a more evenly distribution of the bins a logarithmic transformation seems in order. Furthermore, removal of the cases with missings would get rid of the warning. 

---

b. __Run the code again, but remove cases with missing values with `na.omit()`, set the number of bins to 20, and plot the logarithm of sales using the function `scale_x_log10()`.__

```{r}

```

---

c. __Add another line of code to display 12 facets with a histogram of the log of sales per month.__


```{r}

```

---

d. __Experiment by displaying different themes for the plot.__

```{r}

```

---


# `ggplot2` and `dplyr`

The packages `ggplot2` and `dplyr` are constructed in such a way that they cooperate prefectly. The next exercises are examples of this. The code for combining functions from the two packages is:

```
<DATA> %>% 
   dplyr functions %>% 
   ggplot() +
     plot specifications 
```

Note that the `ggplot()` function no longer requires the `data` argument, because it auromatically takes the result of the pipeline as data argument.

---

a. __With a combination of dplyr functions, remove the cases with missing values, create the new variable `logsales` and `logvolume` with the function `log10()`, and display a scatterplot of these two variables (include the argument `alpha=0.1` to make the points in the plot transparent) . What's the difference in the scale of the x-axis compared to the histograms?__



```{r}

```

---


Suppose you are especially interested in the period 2010-2015.  

b. __Make the same plot as before, but first use a `dplyr` function to select data from 2012-2015. Also give the points from different years another color> There is a difference if you treat `year`  as `numeric` or as `factor` (check it out!), but for now treat year as numeric.__

```{r}

```

---

The colors for year as numeric variable change from ligth to dark blue. You can change the colors with the function `scale_colour_gradient()`

c. __Change the colors  in "yellow" for the first year and "red" for the last year, and check the result.__

```{r}

```

---

The relationships between the two variables seem pretty linear, so we like to add linear regression lines for each year to see if the slopes dfiffer.

d. __Add the linear regression lines for each year (you need to specify year as `factor` now!) to the plot with the function `geom_smooth()`. Specify as method "lm", and suppress the display of the confidence intervals.__


```{r}

```

---

It can be seen that the residuals are evenly distributed over the month, so we may expect the confidence intervals to also fluctuate over the months. To get a better picture of that, we like to inspect separate plots for each year.

e. __Divide the plots over four facets, one for each year, and include the confidence intervals in these plots. What do you notice about the confidence intervals? __


```{r}

```

---

---

END OF LAB.


